Three Algorithms for Merging Hierarchical Navigable Small World Graphs

### 2025-05-21 paper/main.tex
fix affiliations
Apply Sandeep comments to the paper

\affil[1]{HSE Univerity}
\affil[2]{YugaByte LLC}

@@

    \item \textbf{Mapping-based methods}: These techniques transform the original data into representations that allow for efficient search. This category includes Locality-Sensitive Hashing (LSH) \cite{Indyk1998,Gionis1999,Datar2004,Andoni2006,Andoni2015} and Product Quantization approaches \cite{Jegou2011,Ge2013,Norouzi2013}, which encode vectors into compact codes that approximate distances between points.

    \item \textbf{Navigable graph-based methods}: The most recent and currently state-of-the-art approach that constructs graph structures where vertices represent data points and edges connect similar points. These graphs are designed to allow ``greedy-like'' algorithms to perform directed traversal through the data space.
\end{enumerate}

@@

HNSW \cite{hnsw} extends the navigable graphs concept by organizing points into a hierarchy of navigable graphs also named as layers. Each layer contains a progressively smaller subset of data points. The bottommost layer contains all points, while higher layers form a sparse representation of the data space. 
Graphs of all levels glued together form a navigable small world graph similar to NSW  \cite{nsw2011,nsw2012,nsw2014}. An explicit separation set of edges of different lengths to layers helps to reduce the number of computations in search while traversing vertexes with high degrees, also called hubs. A schematic example of HNSW structure presented in Fig. \ref{fig:hnsw}.

@@

\begin{enumerate}
    \item \textbf{Processing Vertex Selection}: Choosing a vertex $v^*$ for which we aim to construct a neighborhood in the merged graph, and determining which original graph it belongs to.

@@

%We provide 3 merge algorithms: "Naive Merge", "IGTM", and "CGTM". We also refer to these algorithm as layer-merge algorithms. "Naive Merge" is a straight forward method, while "IGTM" and "CGTM" are more advanced. Together, these two algorithms constitute the main contribution of the work. The HNSW structure is a sequence of graphs, called layers. We start from description of a general algorithm for merging multi layers structure.
We now present three algorithms for merging HNSW graphs: \textsc{SIGM},  \textsc{NGM}, \textsc{IGTM}, and \textsc{CGTM}. These algorithms operate layer by layer. We first describe the general procedure for merging the multi-layer HNSW structure, followed by the specific logic for merging a single layer (\textit{layer-merge algorithms}).

\subsection{Simple Insertion Graph Merge}
Before we start, we need provide a short descriptopn of simple insertion graph merge (SIGM). It is a default merging strategy that uses an insertion operation to merge graphs. It takes the largest graph and sequentially inserts data from other graph. Actually, it is not merging procedure, but we use it as a basic benchmark.

@@

Note that the merging of different layers (iterations of the loop in Algorithm~\ref{alg:general_merge}) is independent and can potentially be parallelized.

@@

\subsection{Intra Graph Traversal Merge}
The most effort of the \textsc{NGM} algorithm lies in obtaining the set of neighborhood candidates from the other graph utilising \textsc{HNSW-Search} procedure, that every time traverses layer graphs from the top-level down to layer number $\ell$. The number of computations can be reduced if we select the next vertex to process $v^*$ close to the previous one (line 15), instead of randomly choosing it. Thus, for the new $v^*$ the neighborhood candidates will also be close to the previous candidates set. To search these new neighborhood candidates we can use \textsc{LocalSearch} procedure, that traverses the same graph staring from the previous neighborhood candidates set $\mathcal{P}^b$ (lines 11). In line 14 in set $\mathcal{P}^b$ we keep only $M$-closest to $v^*$ candidates.\\
In lines 15,16 we select new processing vertex $v^*$  close to the previous processing vertex $v^*$ that was not already processed.  
To ensure that the new processing vertex $v^*$ is not very far from the previous $v^*$ we bound the size of the results of the \textsc{LocalSearch} procedure controlling by next\_step\_k parameter. Once \textsc{LocalSearch} can't find enough close not processed vertex (if condition in line 17), in line 7 we choice new $v^*$ from the set $\mathcal{V}_{not\_done}$ randomly. \\
After we have processed all vertices from the graph $G^a$. We do the same for the vertices of graph $G^b$ (lines 21-22).

@@

@@

Cross Graph Traversal Merge (\textsc{CGTM}) algorithm is similar to \textsc{IGTM} utilizes \textsc{LocalSearch} procedure to reduce computation efforts. The difference is that \textsc{IGTM} algorithm chooses the next processing vertex $v^*$ from the same graph, while \textsc{CGTM} looks for the new processing vertex $v^*$ in both graphs $G^a$, and $G^b$. 
Thus in line 24 $v^*$ is chosen from the set $\mathcal{C}_{not\_done}$, which is one of the not processed vertex of both graph (line 21). The intuition underlying \textsc{CGTM} algorithm is that when we choose a new processing vertex $v^*$ from both graphs, we reduce the number of times when $v^*$ is chosen randomly, thus minimizing the number of times that we use the more expensive search procedure \textsc{HNSW-Search}.

@@

We evaluated the proposed merge algorithms on the standard ANN benchmark dataset SIFT1M of 1 million 128-dimensional vectors. We divided it into two disjoint subsets of 500k vectors. On those subsets we built two HNSW indices $\mathcal{H}_a$ and $\mathcal{H}_b$ using the following parameters: $M = 16$, $M0$ = 32, ef\_construction = 32.
These indices are then merged using simple insertion strategy \textsc{SIGM} and the three proposed algorithms: \textsc{NGM}, \textsc{IGTM}, and \textsc{CGTM}. We keep the same values of the parameters $m = 16$, and $m0 = 32$ as the target neighborhood size in the merged graph.

@@

Formally, for a given query point \(q\), let \(P_k(q)\) denote the set of \(k\) true nearest neighbors, and let \(A_k(q)\) be the set of top \(k\) points returned by the algorithm. Then, the recall@k is defined as:

@@

The searching quality of the merged graph represented in Fig. ~\ref{fig:search}. The graph plots trade-off between recall and the number of distance computations. As can be seen, \textsc{SIGM} with ef\_construction $\ge$ 32 and \textsc{NGM} with parameter jump\_ef = 20 produce merged graph with slightly better searing quality than other algorithms.

@@

\begin{itemize}
    \item\textsc{IGTM} and \textsc{IGTM}  in our experiment setup  achieve recall better than \textsc{SIGM} with parameter ef\_construciton = 24;
    \item \textsc{NGM} and \textsc{SIGM} perform the most computations, but achieves the highest recall, as it exhaustively reconstructs each neighborhood;
    \item \textsc{CGTM} achieving comparable recall with approximately 60\% fewer computations than \textsc{NGM} and \textsc{SIGM};
    \item \textsc{IGTM} provides the best run-time performance, reducing distance computations about 20\% than \textsc{CGTM}, and about 70\% than \textsc{NGM} and \textsc{SIGM} while achieving comparable recall.
\end{itemize}

@@

\section{Related Works on Graph Merging}
\label{sec:relatedworks}
While working on the paper, we have found that just recently researchers from \textbf{Elastic Search Lab} published a blog post \cite{ElasticHNSW2023} where they proposed fast merge algorithm, which also utilizes information about closeness of object in all graphs. The authors presented a scheme where a subset of vertices $J$ ``join set'' from the smaller graph is selected by greedy heuristic and inserted into the larger graph using standard HNSW insertion procedure. For the remaining vertices from the small graph, their already inserted neighbors and those neighbors' connections in the large graph are used to perform a limited beam search (named FAST-SEARCH-LAYER) to find connection candidates. This approach limits full insertions and accelerates the merging process.

>>>>

\affil[1]{Laboratory LATNA, HSE University}
\affil[2]{Yugabyte, Inc}

@@

    \item \textbf{Mapping-based methods}: These techniques transform the original data into representations that allow for efficient search. This category includes Locality-Sensitive Hashing (LSH) \cite{Indyk1998,Gionis1999,Datar2004,Andoni2006,Andoni2015} and Product Quantization approaches \cite{Jegou2011,Ge2013,Norouzi2013}, which encode vectors into compact codes that approximate the distances between points.

    \item \textbf{Navigable graph-based methods}: The most recent and currently state-of-the-art approach that constructs graph structures where vertices represent data points and edges connect similar points. These graphs are designed to allow ``greedy-like'' algorithms to perform a directed traversal through the data space.
\end{enumerate}

@@

HNSW \cite{hnsw} extends the navigable graphs concept by organizing points in a hierarchy of navigable graphs also named layers. Each layer contains a progressively smaller subset of data points. The bottommost layer contains all points, while the upper layers form a sparse representation of the data space. 
Graphs of all levels glued together form a navigable small world graph similar to NSW  \cite{nsw2011,nsw2012,nsw2014}. An explicit separation set of edges of different lengths to layers helps to reduce the number of computations in search while traversing vertexes with high degrees, also called hubs. A schematic example of HNSW structure presented in Fig. \ref{fig:hnsw}.

@@

\begin{enumerate}
    \item \textbf{Processing Vertex Selection}: Choosing a vertex $v^*$ for which we aim to construct a neighborhood in the merged graph and determining which original graph it belongs to.

@@

%We provide 3 merge algorithms: "Naive Merge", "IGTM", and "CGTM". We also refer to these algorithm as layer-merge algorithms. "Naive Merge" is a straight forward method, while "IGTM" and "CGTM" are more advanced. Together, these two algorithms constitute the main contribution of the work. The HNSW structure is a sequence of graphs, called layers. We start from description of a general algorithm for merging multi layers structure.
We now present three algorithms for merging HNSW graphs: \textsc{NGM}, \textsc{IGTM}, and \textsc{CGTM}. These algorithms operate layer by layer. We first describe the general procedure for merging the multi-layer HNSW structure, followed by the specific logic for merging a single layer (\textit{layer-merge algorithms}).

\subsection{Simple Insertion Graph Merge}
Before we start, we need to provide a short description of simple insertion graph merge (SIGM). It is a default merging strategy that uses an insertion operation to merge graphs. It takes the largest graph and sequentially inserts data from other graph. Actually, it is not a merging procedure, but we use it as a basic benchmark.

@@

% Note that the merging of different layers (iterations of the loop in Algorithm~\ref{alg:general_merge}) is independent and can potentially be parallelized.

@@

\subsection{Intra Graph Traversal Merge}

@@

The most effort of the \textsc{NGM} algorithm lies in obtaining the set of neighborhood candidates from the other graph utilizing the \textsc{HNSW-Search} procedure, which every time traverses the layer graphs from the top level down to the layer number $\ell$. The number of computations can be reduced if we select the next vertex to process $v^*$ close to the previous one (line 15 of the algorithm ~\ref{alg:IGTM}), instead of randomly choosing it. Thus, for the new $v^*$ the neighborhood candidates will also be close to the previous candidates set. To search for these new neighborhood candidates we can use the \textsc{LocalSearch} procedure which traverses the same graph staring from the previous neighborhood candidates set $\mathcal{P}^b$ (lines 11). In line 14 in the set $\mathcal{P}^b$ we keep only $M$-closest to $v^*$ candidates.\\
In lines 15,16, we select a new processing vertex $v^*$  close to the previous processing vertex $v^*$ that was not already processed.  
To ensure that the new processing vertex $v^*$ is not very far from the previous $v^*$ we bound the size of the results of the \textsc{LocalSearch} procedure controlling by next\_step\_k parameter. Once \textsc{LocalSearch} cannot find enough close unprocessed vertex (if condition in line 17), in line 7 we choose a new $v^*$ from the set $\mathcal{V}_{not\_done}$ randomly.\\
After we have processed all vertices from the graph $G^a$. We do the same for the vertices of graph $G^b$ (lines 21-22).

@@

Cross Graph Traversal Merge (\textsc{CGTM}) algorithm is similar to \textsc{IGTM} utilizes the \textsc{LocalSearch} procedure to reduce computation efforts. The difference is that the \textsc{IGTM} algorithm chooses the next processing vertex $v^*$ from the same graph, while \textsc{CGTM} looks for the new processing vertex $v^*$ in both graphs $G^a$, and $G^b$. 
Thus in line 24 $v^*$ is chosen from the set $\mathcal{C}_{not\_done}$, which is one of the not processed vertex of both graph (line 21). The intuition underlying \textsc{CGTM} algorithm is that when we choose a new processing vertex $v^*$ from both graphs, we reduce the number of times when $v^*$ is chosen randomly, thus minimizing the number of times that we use the more expensive search procedure \textsc{HNSW-Search}.

@@

We evaluated the proposed merge algorithms on the standard ANN benchmark dataset SIFT1M of 1 million 128-dimensional vectors. We divided it into two disjoint subsets of 500k vectors. In these subsets, we built two HNSW indices $\mathcal{H}_a$ and $\mathcal{H}_b$ using the following parameters: $M = 16$, $M0$ = 32, ef\_construction = 32.
Then these indices were merged using a simple insertion strategy \textsc{SIGM} and the three proposed algorithms: \textsc{NGM}, \textsc{IGTM}, and \textsc{CGTM}. We keep the same values of the parameters $m = 16$ and $m0 = 32$ as the target neighborhood size in the merged graph.

@@

Formally, for a given query point \(q\), let \(P_k(q)\) denote the set of \(k\) true nearest neighbors, and let \(A_k(q)\) be the set of the top \(k\) points returned by the algorithm. Then, recall@k is defined as:

@@

The searching quality of the merged graph is represented in Fig. ~\ref{fig:search}. The graph plots trade-off between recall and the number of distance computations. As can be seen, \textsc{SIGM} with ef\_construction $\ge$ 32 and \textsc{NGM} with parameter jump\_ef = 20 produce merged graph with slightly better searing quality than other algorithms.

@@

\begin{itemize}
    \item\textsc{IGTM} and \textsc{CGTM} in our experiment setup  achieve recall better than \textsc{SIGM} with parameter ef\_construciton = 24;
    \item \textsc{NGM} and \textsc{SIGM} perform the most computations, but achieves the highest recall, as it exhaustively reconstructs each neighborhood;
    \item \textsc{CGTM} achieving comparable recall with approximately 60\% fewer computations than \textsc{NGM} and \textsc{SIGM};
    \item \textsc{IGTM} provides the best run-time performance, reducing distance computations about 20\% than \textsc{CGTM}, and about 70\% than \textsc{NGM} and \textsc{SIGM} while achieving comparable recall.
\end{itemize}

@@

\section{Related Works on Graph Merging}
\label{sec:relatedworks}
While working on the paper, we have found that just recently researchers from \textbf{Elastic Search Lab} published a blog post \cite{ElasticHNSW2023} where they proposed a fast merge algorithm, which also utilizes information about closeness of object in all graphs. The authors presented a scheme where a subset of vertices $J$ ``join set'' from the smaller graph is selected by greedy heuristic and inserted into the larger graph using standard HNSW insertion procedure. For the remaining vertices from the small graph, their already inserted neighbors and those neighbors' connections in the large graph are used to perform a limited beam search (named FAST-SEARCH-LAYER) to find connection candidates. This approach limits full insertions and accelerates the merging process.
